---
title: "May Institute 2020"
author: "Laurent Gatto"
date: "7 May 2020"
output:
  BiocStyle::html_document:
    toc_float: true
---



```{r style, echo = FALSE, results = 'asis'}
BiocStyle::markdown()
```

# Introduction

These are the live notes from my talk at the (remote) May Institute
2020. I will present material from 

   MSnbase, efficient and elegant R-based processing and visualisation
   of raw mass spectrometry data. Laurent Gatto, Sebastian Gibb,
   Johannes Rainer **bioRxiv** 2020.04.29.067868; doi:
   https://doi.org/10.1101/2020.04.29.067868

See also

- The `MSnbase` home page: http://lgatto.github.io/MSnbase/

- [MSnbase - an R/Bioconductor package for isobaric tagged mass
   spectrometry data visualisation, processing and
   quantitation](https://academic.oup.com/bioinformatics/article/28/2/288/199094). Gatto
   L. and Lilley K.S. Bioinformatics, 28(2), 288-289, 2012.

Acknowledgement: Johannes Rainer and Sebastian Gibb

Questions about your data/analyses/issues:

- Bioconductor support site: https://support.bioconductor.org/
- MSnbase issues on GitHub: https://github.com/lgatto/MSnbase


# On-disk backend


```{r}
f <- msdata::proteomics(pattern = "201412", full.names = TRUE)
basename(f)
```

```{r, message = FALSE}
library(MSnbase)
library(magrittr)
```

Reading on-disk is faster. See **Figure 1a** in the pre-print.

```{r}
system.time(x_mem <- readMSData(f, mode = "inMemory"))
system.time(x_dsk <- readMSData(f, mode = "onDisk"))
```

The `MSnExp` class

```{r}
x_dsk

fData(x_dsk)
fvarLabels(x_dsk)

sp <- x_dsk[[1]]
sp
head(mz(sp))
head(intensity(sp))
plot(sp)
```

The on-disk data contains all spectra

```{r}
length(x_dsk)
length(x_mem)
table(msLevel(x_dsk))
```

On-disk data uses less RAM. See **Figure 1b** in the pre-print.

```{r}
pryr::object_size(x_mem)
pryr::object_size(x_dsk)
```

```{r}
x_dsk2 <- filterMsLevel(x_dsk, 2L)
```

Access to raw data is faster when already in RAM. See **Figure 1d** in
pre-print.

```{r}
system.time(x_dsk[[1]])
system.time(x_mem[[1]])
```

**Filtering**: retention time between 1000 and 3000 seconds

```{r}
chr1 <- chromatogram(x_dsk)
chr1
plot(chr1)
```

**Filtering**: keep TMT6 MZ range

```{r}
plot(x_dsk2[[1227]], reporters = TMT6, full = TRUE)
TMT6
range(mz(TMT6))
```

**Filtering**: keep 50% spectra with highest precursor ion intensity

```{r}
median(precursorIntensity(x_dsk2))
```


```{r}
filter_ms <- function(x) {
   x <- x %>% 
      filterRt(c(1000, 3000)) %>% 
      filterMz(c(120, 135))
   x[precursorIntensity(x) > 11000000, ]
}
```

Filtering on-disk is much faster thank to lazy evaluation. See
**Figure 1c** in the pre-print.

```{r, warning = FALSE}
x_filt <- filter_ms(x_dsk2)
filter_ms(x_mem)
```


```{r}
x_filt[[1227]]
```


```{r, eval=FALSE}
writeMSData
```


# Prototyping: MSnbaseBoxCar

- Reference: [BoxCar acquisition method enables single-shot proteomics
  at a depth of 10,000 proteins in 100
  minutes](https://www.nature.com/articles/s41592-018-0003-5).

- Package home page: http://lgatto.github.io/MSnbaseBoxCar/


```{r}
library(MSnbaseBoxCar)

bcf <- 
   dir(system.file("extdata", package = "MSnbaseBoxCar"),
    pattern = "boxcar.mzML", 
    full.names = TRUE)
```

Read the data

```{r}
bc <- readMSData(bcf, mode = "onDisk") 
bc
```

First full spectrum and 3 first BoxCar spectra

```{r}
p <- plot(bc[1:4])
```

Identify sets of BoxCar spectra

```{r}
fData(bc)$filterString

tmp1 <- bc_groups(bc)
fData(tmp1)$bc_groups
```

Remove non-BoxCar spectra

```{r}
tmp2 <- filterBoxCar(tmp1)
fData(tmp2)$bc_groups
tmp2
```


```{r, warning = FALSE}
bc_plotly(tmp2[1:3])
```

Remove peaks outside of BoxCar segments

```{r, warning = FALSE}
tmp3 <- bc_zero_out_box(tmp2)
bc_plotly(tmp3[1:3])
```

Combine BoxCar spectra and segments into a new full scan

```{r}
res <- combineSpectra(tmp3, 
               fcol = "bc_groups",
               method = boxcarCombine)
res
```

```{r, eval=FALSE}
writeMSData(res, file = "boxcar2.mzML")
```

Putting it all together with `%>%`

```{r}
bc2 <- bc %>% 
   bc_groups() %>% 
   filterBoxCar() %>% 
   bc_zero_out_box() %>% 
   combineSpectra(fcol = "bc_groups",
                  method = boxcarCombine)

```

